Using an SMT engine to generate Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-03-19 2017-Xudong-SMT/ActaInformatica.tex
OT examples, and Xudong TODOs

\begin{example}[An open transition]\TODO{change for Enable example}
  Transition $OT_1$ of the pNet $EnableCompl$ is generated by
  combination of the first vector of its root node, with
  the second vector of its $2^{nd}$ node. Its generated predicate is
  trivialy true. The global states shows that controler $C_3$ has
    changed from state 0 to 1.

    \TODO{add another example to illustrate assignements/Post}

  \label{OT:example-1}
        \begin{mathpar}
	  OT_1  =
          \inferrule*[myfraction=\reddottedrule]
	{\{\xrightarrow{\delta(x3)}_P ~~
          \xrightarrow{acc(x3)}_Q \} ~~
          [\forall z. acc(x3) \neq \delta(z)]
        }
	{\ostate{0,0} \xrightarrow {\underline{\delta(x3)}} \ostate{1,0}}
	\end{mathpar}
\end{example}

@@

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require A pNet $P$ (can be a pLTS, but not a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
          \State Add $\mathit{OT}$ to $L$;
          \State Let $S'$ be the target state of $\mathit{OT}$
          Add $S'$ to the unexplored states if necessary:
	  \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
	  \EndFor
\EndWhile
\State \ERIC{Prune the unsatisfiable open transitions in $L$ using the SMT
solver; only keep reachable global states;}
\State \Return $\mathit{OA}=(\mathit{InitState}(P),L)$;

@@

\ERIC{Before the SAT checking step, it may contain symbolic transitions that
do not have valid ground instanciations. This is not logicaly
incorrect, but is naturaly not an optimal representation. So we prune
these UNSAT transitions: at the root of the pNet, the predicate of each OT is translated
into SMTlib assertions, and checked for satisfiability.
The final open automaton includes all satisfiable OTs, and the set of
reachable global states.}
%By construction, due to the
%satisfiability test in line 6, it only produces the set of states reachable
%through a sequence of satisfiable open transitions.

@@

\paragraph{Filtering.}
While matching a vector with a combination tuple, \emph{Matching}
tries to filter out some incompatibilities; there may be several
reasons why the matching would fail: 
\begin{itemize}
  \item if some subnet is marked as inactive in the vector, and the
    chosen combination has an active behaviour at this position,
  \item if some subnet action expression in the vector does not match (by
    pattern-matching) the corresponding action expression in $C$,
  \item or if the whole set of active subnet actions in the vector
    cannot be matched (by unification) with the corresponding action expressions
      on the tuple $C$.
\end {itemize}

@@

In a pLTS, there may be several incoming transitions 
that assign potentially different values to a state variable.
To handle such cases, the algorithm manages, for each pLTS state, a
list of expressions collected from the assignments of each state
variable. %% $v$.
%% The list we used contains triples <$v$, $S$, $AssignRH$> 
%% where $v$ is the variable in pLTS, $S$ is its owner state and
%% $AssignRH$ is a list of expressions over other variables in the right
%% hand side of assignments of $v$.
% As a pragmatic extension to the formal definition, we also manage
% ``global variables'', defined in all states of the pLTS.
For a global state in the open automaton, the set of state
variables (which may be used in a transition) %% will simply be
is the disjoint union
of sets of state variables of the individual pLTS states constituting this
global state.
%There are some assignment on the variables in each of states of the pLTS (state variables), or just as global states of it (global variables). There is usually not only one assignment for a state variable, miss of them may drop the unsatisfiable results. In order to avoid such mistakes, 
\TODO{these assignment sets collect potential assignments from all OTs
  in the generated automaton. If we minimize the generated state
  space, we will get less assignements, and potentialy also get better
  minimization... }

@@

\ERIC{But there is more to this. As mentionned above, we collect all
possible (symbolic) assignments of state variables, in the
computed open automaton. By avoiding building the OTs coming from
unreachable states, we also avoid adding the corresponding assignments
to the variables of their target states. The set of possible
assignments is an over approximation of possible values of the
variables of such state, so narrowing it means we get a better
approximation, and potentialy more informations, that can be used to
the SAT check of subsequent transitions.}
\TODO{Beware, I think this is dangerous... Xudong, do we really use
  assignment information in SAT checking ? In the ``smart'' version, I
  think we must NOT use it, because we do not yet have collected all
  possible assignments.}

@@

\TODO{Candidate theorems:
  - All reachable states from algorithm 1 are kept by Algorithm 3\\
  - Algorithm 3 computes an open automaton that is minimal with
  respect of SAT checking decidability.\\
  I have problem with minimality... Because of the ``two phases'' SAT
  checking that we have now, because we cannot use incomplete sets of
  variable assignments during the ``smart'' phase, and use them later,
minimality becoomes trivialy equivalent in the original and smart algorithms...}

@@

                (with 2 client holes)&
                2496 ms 				&
                844 ms  
		\\\hline\specialrule{0em}{1pt}{1pt}
		FailureTimer    			&
                12/364 OTs           &
                12/167 OTs
                \\\specialrule{0em}{1pt}{1pt}

  		&
                1080 ms 				&
                563 ms  
		\\\hline\specialrule{0em}{1pt}{1pt}
		FailureTimer X 2    			&
                25/8964 OTs           &
                25/1255 OTs
                \\\specialrule{0em}{1pt}{1pt}

>>>>

\begin{example}[Open transitions]
  Transition $OT_1$ of the pNet $EnableCompl$ is generated by
  combination of the first vector of its root node, with
  the second vector of its $2^{nd}$ node. Its generated predicate is
  trivialy true. The global states shows that controler $C_3$ has
    changed from state 0 to 1.

  \label{OT:example-1}
        \begin{mathpar}
	  OT_1  =
          \inferrule*[myfraction=\reddottedrule]
	{\{\xrightarrow{\delta(x3)}_P ~~
          \xrightarrow{acc(x3)}_Q \} ~~
          [\forall z. acc(x3) \neq \delta(z)]
        }
	{\ostate{0,0} \xrightarrow {\underline{\delta(x3)}} \ostate{1,0}}
	\end{mathpar}

If we use the data-oriented pNet Enable2 from Figure \ref{schema:enable-pnets} to building a similar
``P>>(Q>>R)'' expression, the corresponding open transition would be:

  \label{OT:example-1}
        \begin{mathpar}
	  OT_2  =
          \inferrule*[myfraction=\reddottedrule]
	{\{\xrightarrow{\delta(x3)}_P ~~
          \xrightarrow{acc(x3)}_Q \} ~~
          [v_1=0 \land v_2 = 0 \land \forall z. acc(x3) \neq
            \delta(z)] ~~ \{v1:=1\}
        }
	{\ostate{0,0} \xrightarrow {\underline{\delta(x3)}} \ostate{0,0}}
	\end{mathpar}
        where $v_1$ and $v_2$ are respectively the state variables of
        the first and second Enable2 pNet. Indeed the global
        state space of this version is reduced to $\ostate{0,0}$, and
        the state change is replaced by the assignment ``$v_1:=1$''.

\end{example}

@@

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require A pNet $P$ (can be a pLTS, but not a hole)
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(P)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; remove $S$ from $U$, add $S$ to $E$;
        %\qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OTs} = \mathit{MakeTransitions}(P, S)$;
	\For {each $\mathit{OT} \in \mathit{OTs}$}
          \State Add $\mathit{OT}$ to $L$;
          \State Let $S'$ be the target state of $\mathit{OT}$
          Add $S'$ to the unexplored states if necessary:
	  \IfThen{($S' \not\in U \cup E$)}{Add $S'$ into $U$;}
	  \EndFor
\EndWhile
\State \TODO{Xudong:Prune the unsatisfiable open transitions in $L$ using the SMT
  solver;
   and add a call to ``compute the reachable sub-automaton'';}
\State \Return $\mathit{OA}=(\mathit{InitState}(P),L)$;

@@

\ERIC{Before the SAT checking step, it may contain symbolic transitions that
do not have valid ground instanciations. This is not logicaly
incorrect, but is naturaly not an optimal representation. So we prune
these UNSAT transitions: at the end of the open automaton generation,
the predicate of each OT is translated 
into SMTlib assertions, and checked for satisfiability.\\
Then we compute the set of reachable transitions and states to build
the final open automaton.}
%By construction, due to the
%satisfiability test in line 6, it only produces the set of states reachable
%through a sequence of satisfiable open transitions.

@@

\paragraph{Filtering.}
While matching a vector with a combination tuple, \emph{Matching}
tries to filter out some incompatibilities; there may be several
reasons why the matching would fail: 
\begin{itemize}
  \item if some subnet is marked as inactive in the vector, and the
    chosen combination has an active behaviour at this position, or vice-versa;
  \item if some subnet action expression in the vector does not match (by
    pattern-matching) the corresponding action expression in $C$;
  \item or if the whole set of active subnet actions in the vector
    cannot be matched (by unification) with the corresponding action expressions
      on the tuple $C$ (this is strictly stronger than the
      previous filter). 
\end {itemize}

@@

In a pLTS, there may be several incoming transitions 
that assign potentially different (symbolic) values to a state variable.
To handle such cases, the algorithm manages, for each pLTS state, a
list of expressions collected from the assignments of each state
variable. %% $v$.
%% The list we used contains triples <$v$, $S$, $AssignRH$> 
%% where $v$ is the variable in pLTS, $S$ is its owner state and
%% $AssignRH$ is a list of expressions over other variables in the right
%% hand side of assignments of $v$.
% As a pragmatic extension to the formal definition, we also manage
% ``global variables'', defined in all states of the pLTS.
For a global state in the open automaton, the set of state
variables (which may be used in an outgoing transition) %% will simply be
is the disjoint union
of sets of state variables of the individual pLTS states constituting this
global state.
\TODO{Xudong: speciy more precisely what is done, and at which point
  in the MakeTransitions algo}
%There are some assignment on the variables in each of states of the pLTS (state variables), or just as global states of it (global variables). There is usually not only one assignment for a state variable, miss of them may drop the unsatisfiable results. In order to avoid such mistakes, 
\TODO{Eric: these sets of assignment collect potential assignments from all OTs
  in the generated automaton. If we prune the generated state
  space, we will get less assignements, and potentialy also get better
  minimization... May be move this to ``smart'' section}

@@

\ERIC{But there is more to this. As mentionned above, we collect all
possible (symbolic) assignments of state variables, in the
computed open automaton. By avoiding building the OTs coming from
unreachable states, we also avoid adding the corresponding assignments
to the variables of their target states. The set of possible
assignments is an over approximation of possible values of the
variables of such state, so narrowing it means we get a better
approximation, and potentialy more informations, that can be used to
the SAT check of subsequent transitions. But we \emph{cannot} use this
assignment information in the ``on-the-fly'' SAT checking, because
they cannot be complete before exploring (at least) all the reachable
state space. So the smart algorithm performs in to steps: during the
first step, the SAT check is done without using any assignment set
information. Then after completion of this step, we perform another
SAT check of all transitions from step1, but this time using the
assignment information. This may reduce the reachable state space. If
we want to be sure the obtained automaton is minimal, we should
iterate this step until a fix-point is reached. But in practical cases, very
little will be gained, and leaving some UNSAT transitions in the
symbolic automaton is still logicaly correct.}

@@

\TODO{Candidate theorems: Xudong: state the theorems, and sketch the proofs}

\ERIC{First the good case: if we have no assignments to state variables in
the system, then we can directly compute the reachable part of the
automaton using the on-the-fly algorithm.}

\begin{theorem}[Correctness without assignments]
  \label{Th:finiteness}
If a pNet contains no assignment to state variables, then the smart
algorithm, without final reachability computation, computes exactly
the same open automaton as algorithm 1.
  \end{theorem}

\TODO{Xudong: proof}

\ERIC{But in presence of assignments, we need to collect all of them
  before using the assignment information to refine the SAT
  checking. Still the smart strategy will explore a reduced state
  space, but not the minimal one (discussion later)}

\begin{theorem}[Correctness without assignments]
  \label{Th:finiteness}
In the general case, then the smart
algorithm, with a final reachability computation, computes exactly
the same open automaton as algorithm 1.
\end{theorem}

\TODO{Xudong: proof}

\TODO{Eric : provide counter-example for smart with unsing incomplete
  assignment sets}

@@

                (with 2 client holes)&
                2496 ms 				&
                844 ms  
		\\\hline\specialrule{0em}{1pt}{1pt}
		FailureTimer    			&
                12/396 OTs           &
                12/192 OTs
                \\\specialrule{0em}{1pt}{1pt}

  		&
                1220 ms 				&
                687 ms  
		\\\hline\specialrule{0em}{1pt}{1pt}
		FailureTimer X 2    			&
                25/8964 OTs           &
                25/1255 OTs
                \\\specialrule{0em}{1pt}{1pt}
